Value-Suppressing Uncertainty Maps

### 2017-09-16 paper/main.tex
Dealt with all JH edits except for the difficult one.

Uncertainty is an inescapable component of collecting, presenting, and using data. A common goal in the communication of uncertainty is promoting \emph{uncertainty-aware decisions}: the audience should be aware of the risks and rewards of certain decisions, modulate their confidence in their conclusions, and perhaps refrain from making a decision at all if there is too much uncertainty.  A way that designers can contribute to this goal is by ensuring that uncertainty information is \emph{well-integrated} with the rest of the data. That is, it should be difficult to discount or ignore the uncertainty in a dataset.

@@

We describe the motivations for VSUPs, provide examples of their utility for decision-making under uncertainty, and assess VSUPs in a crowdsourced experiment. Our results indicate that VSUPs create close integration between uncertainty and data, promoting rapid but cautious decision-making.

@@

There are many methods for visualizing uncertainty. MacEachren \ea~\cite{maceachren1992visualizing} present a theoretical survey of uncertainty visualization. Relevant to our domain of heatmaps and thematic maps \textbf{JH: where did this come from? There is no mention of focusing on this domain in the intro. The whole carto/GIS stuff seems to come out of left field.} is work in GIS and cartography; for a survey of uncertainty visualization techniques in GIS, see Kinkeldey \ea~\cite{kinkeldey2014assess,kinkeldey2017evaluating}. Yet, despite a wide variety of methods, to our knowledge, there are no established standards for visualizing uncertainty in heatmaps.

@@

\subsection{Color Maps}

For spatial visualizations like choropleth maps, heatmaps, and treemaps, visual channels such as position and length are reserved for data variables other than value (such as geographic location or relative size). In these situations, data value is often encoded using color. Colors in univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Different choices of color maps can highlight different features of the data, and should be chosen with care~\cite{rogowitz1996not}.

A recurring design consideration when constructing color maps is whether to quantize data into a discrete set of colors, or to encode data using a continuous mapping. While continuous color maps afford greater fidelity in presenting values~\cite{muller1979perception}, non-linearity in human color perception introduces errors in extracting numeric values from continuous colors~\cite{borland2007rainbow}. Quantizing a color map is therefore an exercise in balancing \emph{perceptual} error and \emph{quantization} error~\cite{dobson1973choropleth}. Discrete maps offer finer control over this balance, which can result in better performance in tasks involving heatmaps~\cite{padilla2017evaluating}. As such, we limit our discussion to color maps with discrete sets of colors.

@@

Bivariate maps can, in principle, be constructed from the combination of any two visual variables (such as shape and color, or size and texture). For example Ware's ``textons''~\cite{ware2009quantitative} overlay glyphs on top of regions to simultaneously encode two quantitative values. However, in this work, our focus is on heatmaps and other sorts of thematic maps. In visualizations like heatmaps, common visual channels such as position and size are already in use to encode other information (such as spatial location and extent), and color is reserved for value \textbf{JH: this is a bit redundant with the previous subsection on color maps. It feels a bit artificial having these two sections separate. Color is a subset of bivariate map options... and we start talking about it again in the next paragraph.}. Bivariate maps in these settings therefore typically rely on a visual channel related to color for encoding a secondary variable, such as opacity (alpha)~\cite{roth2010value} or pixel noise~\cite{lucchesi2017visualizing}.

Due to the perceptual integrality of visual channels~\cite{garner1970integrality, ware2012information}, \textbf{JH: We mention integrality but never give a brief definition or explanation. I wouldn't assume that all CHI readers are familiar with this.} the construction of color-based bivariate maps is difficult. Viewers need to be able to independently assess the variables of interest \textbf{JH: They do? I thought the whole point of VSUPs is to confound indepenent assessment of value!}, but the resulting bivariate map should also be expressive and easy to interpret. Bivariate maps with these constraints are often limited to relatively few categories (say, a 4x4 matrix as in \figref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. For VSUPs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analysis tasks, rather than uniformly.

In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for a general audience to interpret~\cite{wainer1980empirical} \textbf{JH: It seems there is a lot ``hiding'' behind this citation. In what ways is it hard? In what ways do people fail? Any more flavor here, even if very brief? And might this motivate something akin to VSUPs?}.

\textbf{JH: This would be a natural place to put some connective prose that transitions to VSUPs...}

@@

Correll \ea~\cite{correll2015layercake,correll2011visualizing} use a precursor of VSUPs in their LayerCake genomics visualization tool, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into a ``confidence fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection to uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both cases, the property of aliasing is \emph{ad hoc}, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUPs makes the choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

\textbf{JH: The precending paragraph might be better placed in related work!}

@@

%\performanceFig

There are multiple choices that designers must make before creating a VSUP. Specifically, they must choose (1) which visual channels map to value and uncertainty and (2) the structure of the underlying VSUP tree.

Often, the graphical conventions of a chart reserve variables such as position and area for other uses, and determine the visual channel to be used for value. In heatmaps, for instance, value is regularly encoded using color, while vertical and horizontal position are reserved for location. \textbf{JH: This is the third time we've pointed this out...} In this paper, we primarily deal with the case where value is encoded using color. While there are some similar conventions for representing uncertainty, generally designers have more freedom in choosing a visual channel for uncertainty~\cite{maceachren1992visualizing}. \textbf{This feels like we are revisiting introductory and related work material again!} We recommend channels with both a strong semiotic connection to uncertainty as well as a relatively large number of perceptually distinguishable levels. Here, we focus on value/saturation (or, equivalently, transparency against a white background). Correll \& Gleicher~\cite{correll2013error} provide evidence that even audiences without statistical training can successfully interpret uncertainty information encoded in these channels.

\textbf{We should say something about the choice of value color map. Generally speaking, single-hue luminance ramps must be used with caution if they ramp too strongly towards white. This will be confused with manipulations of opacity or saturation. Viridis might be attractive due to its avoidance of desaturated colors.}

%TODO Add in discussion of tree structures, including figure showing that VSUPs have colors that are further apart than similar 2d bivariate maps. Also worth discussing discretization issues here. Make new performance fig.

@@

\section{Evaluation}

\conditionFig

@@

We limited our population to Turkers from the United States, with a prior task approval rate of at least 90\%. As the experimental tasks required multi-hue color perception, we presented participants with a set of Ishihara plates~\cite{hardy1945tests} as a pretest, and excluded participants who either misidentified the values in the plates, or who self-reported as having a color vision deficiency (CVD) in the post-test. Based on piloting, we paid participants \$2 dollars for participation, for a target rate of \$8/hour. After completion of the main tasks, we solicited demographic information, including a risk aversion assay from Mandrik \& Bao~\cite{mandrik2005exploring}. We recruited a total of $48$ participants for this experiment ($24$ for each experiment): $N$ female, $N$ male, and $N$ who declined to state ($M_{\text{age}}$= $N$, $SD_{\text{age}}$ = $N$). Participants from the first experiment were excluded from participating in the second.

@@

Our results from the prior study indicate indicate that discrete, non-juxtaposed maps outperformed the other bivariate maps we selected, so we limited our study to only 4 types: square and arc bivariate maps. These in turn were either VSUPs or traditional bivariate maps, for a 2 (square or arc legend) x 2 (VSUP or standard) factorial design, with 6 replications, for a total of 24 stimuli for each of our 24 participants. \figref{fig:conditions} shows these factor levels. Prior to the main task, we included a short replication of the identification task from the prior experiment (with 12 stimuli) for training purposes.

@@

We had two hypotheses, stemming from our belief that VSUPs promote better \emph{integration} between uncertainty and value information, and also encourage \emph{caution} by highlighting the ambiguity or untrustworthiness introduced by uncertain data. In particular:
\begin{enumerate}
	\item Participants would choose targets with \textbf{lower uncertainty} in the prediction task when using a VSUP. 
	\item This would result in a tradeoff where they would also choose targets with \textbf{lower expected value, but lower variance in expected value} when using a VSUP.
\end{enumerate}

@@

@@

\subsection{Limitations and Future Work}
% When should you use these things?
In general, VSUPs excel in scenarios where we wish to make firm guarantees about the discriminability of different marks in a chart, but must also represent, in as much detail as possible, two, potentially orthogonal variables of interest. VSUPs have improved data resolution over other bivariate maps by assuming that uncertain values ought to have less visual impact that highly certain values. VSUPs therefore act both as a kind of filtering device as well as a bivariate representation. This assumption holds in scenarios like the virology dataset (\figref{fig:viralVsum}), where the expectation is that highly uncertain data is unreliable or should otherwise be downweighted in analysis. If the analyst has a different interpretation of uncertainty, or wishes to quickly and orthogonally analyze the distributions of uncertainty and value, other strategies, such as juxtaposed maps, may be more appropriate. VSUPs are designed for the \emph{integration} of value and uncertainty. Designers should take care in considering when and how this integration is desirable.

>>>>

@@

We describe the motivations for VSUPs, provide examples of their utility for decision-making under uncertainty, and assess VSUPs in a crowdsourced experiment. Our results indicate that VSUPs create close integration between uncertainty and data, resulting in better performance at data tasks with uncertainty components.

@@

%There are many methods for visualizing uncertainty. MacEachren \ea~\cite{maceachren1992visualizing} present a theoretical survey of uncertainty visualization. Relevant to our domain of heatmaps and thematic maps \textbf{JH: where did this come from? There is no mention of focusing on this domain in the intro. The whole carto/GIS stuff seems to come out of left field.} is work in GIS and cartography; for a survey of uncertainty visualization techniques in GIS, see Kinkeldey \ea~\cite{kinkeldey2014assess,kinkeldey2017evaluating}. Yet, despite a wide variety of methods, to our knowledge, there are no established standards for visualizing uncertainty in heatmaps.

@@

@@

Bivariate maps can, in principle, be constructed from the combination of any two visual variables (such as shape and color, or size and texture). For example Ware's ``textons''~\cite{ware2009quantitative} overlay glyphs on top of regions to simultaneously encode two quantitative values. However, for spatial visualizations like choropleth maps, heatmaps, and treemaps, visual channels such as position and length are reserved for data variables other than value (such as geographic location or relative size). In these situations, data value is often encoded using color. Bivariate maps in these settings therefore typically rely on a visual channel related to color for encoding a secondary variable, such as opacity~\cite{roth2010value} or pixel noise~\cite{lucchesi2017visualizing}.

Colors in univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Different choices of color maps can highlight different features of the data, and should be chosen with care~\cite{rogowitz1996not}. These principles extend to bivariate color maps, with the additional consideration that our perception of color channels lacks \emph{orthogonality}. That is, while we can perceive e.g., the height of a bar in a bar chart independently of its width, a particular property of a color (e.g., its redness, saturation, transparency, etc.) affect how its other properties are perceived \cite{ware2012information,garner1970integrality}. Many color channels such as hue and saturation are therefore perceptually \emph{integral}, which complicates their use in bivariate maps.

Another recurring design consideration when constructing color maps is whether to quantize data into a discrete set of colors, or to encode data using a continuous mapping. While continuous color maps afford greater fidelity in presenting values~\cite{muller1979perception}, non-linearity in human color perception introduces errors in extracting numeric values from continuous colors~\cite{borland2007rainbow}. Quantizing a color map is therefore an exercise in balancing \emph{perceptual} error and \emph{quantization} error~\cite{dobson1973choropleth}. Discrete maps offer finer control over this balance, which can result in better performance in tasks involving heatmaps~\cite{padilla2017evaluating}.

In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for a general audience to interpret: due to their additional complexity, Wainer \& Francolini~\cite{wainer1980empirical} reported high levels of error for participants even for ``elementary'' graph reading tasks, and additionally that bivariate legends were more difficult to memorize and internalize than their univariate equivalents. Therefore, bivariate maps in practice are often limited to a small set of output colors (say, a 4x4 matrix as in \figref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. For VSUPs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analysis tasks, rather than uniformly.

Given that bivariate maps, for reasons of either practicality or perceptual fidelity, have only a limited budget of outputs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analysis tasks, rather than uniformly. Our Value-Suppressing Uncertainty Palettes apply this insight to data with uncertainty, where, frequently, uncertain information ought to have less valence on the decision-making process, or at least less prominence in the analysis, than certain information.

Correll \ea~\cite{correll2015layercake,correll2011visualizing} use a precursor of VSUPs in their LayerCake genomics visualization tool, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into a ``confidence fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values through perceptual integrality. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection to uncertainty, such as saturation, value, blur, or size, have similar deleterious effects on the disambiguation of colors and shapes. In both cases, the property of aliasing is \emph{ad hoc}, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUPs makes the choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

@@

Since the visual channel used to encode uncertainty (e.g., saturation, lightness, transparency) often reduces the ability of people to distinguish colors, VSUPs have the added benefit of creating bivariate palettes that are perceptually easier to distinguish, by reducing color resolution in precisely the areas where color disambiguation would be most difficult anyway. For instance, the two closest colors in the VSUP in \figref{fig:example} are 1.4 units farther apart in CIELAB color space.

@@

There are multiple choices that designers must make before creating a VSUP. Specifically, they must choose (1) which visual channels map to value (1), uncertainty (2), and the structure of the underlying VSUP tree (3).

When selecting a visual channel to represent uncertainty, we recommend channels with both a strong semiotic connection to uncertainty (such as those recommended by MacEachren et al.~\cite{maceachren2012visual}) as well as a relatively large number of perceptually distinguishable levels (such as those evaluated by Boukhelifa et al. \cite{boukhelifa2012evaluating}). Here, we focus on a combination of increasing color value and decreasing saturation. Correll \& Gleicher~\cite{correll2013error} provide evidence that even audiences without statistical training can successfully interpret uncertainty information encoded in these channels.

When selecting a visual channel to represent value, the designer should consider the perceptual integrality of the value and uncertainty visual channels. For instance, we use the Viridis~\cite{viridis} color map frequently in this paper, as many other standard color ramps (such as sequential ColorBrewer~\cite{harrower2003colorbrewer} ramps), following usual best practices, interpolate in both hue and value~\cite{ware1988color}. This interpolation in value interferes with our uncertainty encoding, introducing ambiguity. Ramps with similar saturation and value among colors, such as Virids, are less ambiguous.

When determining the quantization tree, the design should be careful to avoid having too many output colors. For reasons of memorability and complexity, Wainer\& Francolini~\cite{wainer1980empirical} suggest no more than 16. For a binary quantization tree, this results in a maximum tree depth of 4, as seen in \figref{fig:example}. In general, for tasks that require finer-grade distinctions between uncertainty, the depth of the tree should be increased. For tasks involving finer-grade distinctions between value, the branching factor of the tree should be increased. Beyond these two parameters, non-uniform binning would allow the designer to target particular distributions or important subregions of the data that would otherwise be difficult to visualize in a limited palette, as in Dunn's~\cite{dunn1989dynamic} dynamic bivariate maps. As an additional form of control, prior tools using VSUP-like techniques, such as LayerCake~\cite{correll2015layercake}, allow the analyst to interactively control the size of bins in the tree through dragging on the color legend.

@@

\section{Evaluation}
%TODO update conditions fig
\conditionFig

@@

We limited our population to Turkers from the United States, with a prior task approval rate of at least 90\%. As the experimental tasks required multi-hue color perception, we presented participants with a set of Ishihara plates~\cite{hardy1945tests} as a pretest, and excluded participants who either misidentified the values in the plates, or who self-reported as having a color vision deficiency (CVD) in the post-test. Based on piloting, we paid participants \$2 dollars for participation, for a target rate of \$8/hour. After completion of the main tasks, we solicited demographic information, including a risk aversion assay from Mandrik \& Bao~\cite{mandrik2005exploring}. We recruited a total of $48$ participants for this experiment ($24$ for each experiment): $24$ female, $24$ male, ($M_{\text{age}}$= $34$, $SD_{\text{age}}$ = $9.1$). Participants from the first experiment were excluded from participating in the second.

@@

Our results from the prior study indicate indicate that discrete, non-juxtaposed maps outperformed the other bivariate maps we selected, so we limited our study to only 4 types: square and arc bivariate maps. These in turn were either VSUPs or traditional bivariate maps, for a 2 (square or arc legend) x 2 (VSUP or standard) factorial design, with 6 replications, for a total of 24 stimuli for each of our 24 participants. \figref{fig:conditions} shows these factor levels. Prior to the main task, we included a short replication of the identification task from the prior experiment (with 12 stimuli) for training purposes. Poor performance on this training was used for exclusion purposes. Three people with low accuracy ($6\%$, $6\%$, and $25\%$ accuracy compared to a mean of $70\%$ ) were excluded.

@@

We had two hypotheses, stemming from our belief that VSUPs promote better \emph{integration} between uncertainty and value information, and also encourage \emph{caution} by highlighting the ambiguity or untrustworthiness introduced by uncertain data. In particular:
\begin{enumerate}
	\item Participants would choose targets with \textbf{lower uncertainty} in the prediction task when using a VSUP. 
	\item This would result in a tradeoff where they would also choose targets with \textbf{lower expected value} when using a VSUP.
\end{enumerate}

@@

\subsubsection{Results} 
\taskTwoViolin
Consistent with our first experiment, we found no significant effect of legend shape on either uncertainty ($F(1,61) = 0.01$, $p=0.92$) or value ($(1,61)=3.1$, $p=0.08$) of guesses. This seems to indicate that both wedge- or square-shaped legends promote similar patterns of decision-making. In this paper, and in general, we prefer to employ the wedge-shaped legend for VSUPs, and the square-shaped legend for traditional maps, as it makes the conceptual differences between the two more apparent.

Our results failed to support our first hypothesis. We performed a repeated measures ANOVA on our results to measure the effect of VSUP versus standard quantizations, and square- versus wedge-shaped legends, on average uncertainty in bets. We found no significant effect of quantization scheme on average guess uncertainty ($F(1,61)=0.05$, $p=0.83$). This indicates that there does not appear to be a uniform pattern of risk aversion between the two scale types, as we had surmised.

Our results supported our second hypothesis. A similar ANOVA on average danger value of guesses found that quantization scheme was a significant effect ($F(1,61)=17$, $p<0.01$). A post-hoc t-test confirmed ($t=2.3$, $p=0.02$) that guesses made with VSUPs had significantly higher average danger ($M=0.32$, $SD=0.17$) (and so lower expected value) than traditional maps ($M=0.29$, $SD=0.18$). Since there was no significant difference in average guess uncertainty, and the distribution of danger and uncertainty was identical across conditions, this seems to indicate that participants were using VSUPs moderating their guesses: avoiding highly uncertain regions, but willing to ``gamble'' more on regions of middling certainty. \figref{fig:taskTwoViolin} shows the comparative distribution of uncertainty in participant guesses.

@@

\subsection{Limitations and Future Work}

VSUPs assume that uncertain values ought to have less visual impact that highly certain values. VSUPs therefore act both as a kind of filtering device as well as a bivariate representation. This assumption holds in scenarios like the virology dataset (\figref{fig:viralVsum}), where the expectation is that highly uncertain data is unreliable or should otherwise be downweighted in analysis. If the analyst has a different interpretation of uncertainty, or wishes to quickly and orthogonally analyze the distributions of uncertainty and value, other strategies, such as juxtaposed maps, may be more appropriate. VSUPs are designed for the \emph{integration} of value and uncertainty. Designers should take care in considering when and how this integration is desirable.
